\documentclass[aspectratio=169]{beamer}
\usetheme{metropolis}
\usecolortheme{default}
\usefonttheme{default}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{caption}[numbered]

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}

\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    frame=single,
    backgroundcolor=\color{gray!10},
    commentstyle=\color{green!50!black},
    keywordstyle=\color{blue},
    stringstyle=\color{red},
    showstringspaces=false
}

\title[Forward Selection]{MA2003B - Application of Multivariate Methods in Data Science}
\subtitle{Topic 1.4: Forward Selection}
\author{Dr. Juliho Castillo}
\institute{Tec de Monterrey}
\date{\today}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\begin{frame}{Outline}
  \tableofcontents
\end{frame}

\section{Introduction}
\begin{frame}{What is Forward Selection?}
  \begin{itemize}
    \item Forward selection is a stepwise regression method.
    \item It starts with no predictors and adds them one by one.
    \item At each step, the predictor that improves the model the most is added.
    \item Useful for building parsimonious models.
  \end{itemize}
\end{frame}

\section{Steps in Forward Selection}
\begin{frame}{How Does Forward Selection Work?}
  \begin{itemize}
    \item Start with an empty model.
    \item Evaluate all predictors and add the one with the highest improvement (e.g., lowest p-value).
    \item Repeat until no significant improvement is observed.
    \item Stop when adding predictors no longer improves the model.
  \end{itemize}
\end{frame}

\section{Python Practice}
\begin{frame}{Python Practice: Forward Selection}
  \begin{itemize}
    \item Try the script: \texttt{forward\_selection\_practice.py}
    \item It demonstrates how to implement forward selection using statsmodels.
    \item Follow the comments in the script for guidance.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]{Running the Script}
  \begin{itemize}
    \item Install required libraries if needed:
  \end{itemize}
  \vspace{0.5cm}
  \begin{lstlisting}[language=bash]
pip install numpy pandas statsmodels matplotlib
  \end{lstlisting}
  \vspace{0.5cm}
  \begin{itemize}
    \item Run the script:
  \end{itemize}
  \vspace{0.5cm}
  \begin{lstlisting}[language=bash]
python forward_selection_practice.py
  \end{lstlisting}
\end{frame}

\begin{frame}
  \centering
  \Huge Thank You!
  \vspace{1cm}
  \normalsize Questions?
\end{frame}

\end{document}
